---
title: "1_tutorial_data_visualization"
output:
  html_document: default
  html_notebook: default
---

## Welcome!

Some day in this course you will learn some advanced machine learning techniques, but it is not this day. Today we plot!

We want to start of easy, because we are not sure if everybody is familiar with R,
Today we will share with you some useful resources for R, and teach you the basics of data manipulation and plotting.

It might be that you do not finish all the material of this week, depending on your prior knowledge. However this week there will be no assignment, so we expect you to finish the material of this week's tutorial in your own time.

## Useful Resources

* [Cheat sheets](https://www.rstudio.com/resources/cheatsheets/) -- for this week make sure you have:
  * [RStudio](https://github.com/rstudio/cheatsheets/raw/master/rstudio-ide.pdf)
  * [Data manipulation](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf)
  * [Data visualization](https://github.com/rstudio/cheatsheets/raw/master/data-visualization-2.1.pdf)
* [Online books](https://bookdown.org/home/) -- these books are actually written in R as well. Particularly useful in this course are:
  * [R for data science, written by the legendary Hadley Wickham](http://r4ds.had.co.nz/)
  * [Tidytextmining](http://tidytextmining.com/)
* [Online courses at datacamp.com](http://datacamp.com/) -- You have gotten an invite on your university mail if not please let me know.

## Setting up R for today's session

Install the [tidyverse](https://www.tidyverse.org/), a set of packages which extends basic r with tools you will use all the time.
You can do as follows:

```{r,eval=FALSE}
install.packages("tidyverse")
```

You can then load the tidvyverse as follows:

```{r,results="hide",message=FALSE}
library(tidyverse)
```

**Note how the first statement requires quotation marks, while the latter does not!**

Now we need some data manipulate and plot.
We will use the famous [iris](https://en.wikipedia.org/wiki/Iris_flower_data_set) dataset.
It is already included in R, so you don't need to load it!

## Basic data exploration

There are multiple functions to get a first glance of your data.

Try them all:

```{r,results="hide"}
str(iris)
head(iris)
summary(iris)
View(iris)
```

You see we have data in the sepal and petal width and length for three different flower species.
[What's a petal and sepal?](http://jiangtanghu.com/blog/wp-content/uploads/2011/09/iris.gif)
[what are these species?](https://image.slidesharecdn.com/irisdataanalysiswithr-140801203600-phpapp02/95/iris-data-analysis-example-in-r-3-638.jpg?cb=1406925587)

## Basic plotting

Let's learn a bit more about this data by introducing some plots, a first good step in any data analysis project.

To learn plotting in R we must first learn *the grammar of graphics*.
You can learn more about this in [chapter 3 of R for Data Science](http://r4ds.had.co.nz/data-visualisation.html),
but a look at the cheat sheet might also work fine.

The most important part to remember is that every equation is the combination of:

1. Data -- your full dataset
2. Geometric Function -- the "type" of graph you want
3. Mapping -- the role every column of your data has in the graph
4. Fancy stuff -- scale, theme, facet, etc.

A basic template is then: ggplot(data = <DATA>) + <GEOM_FUNCTION>(mapping=aes(<MAPPINGS>)) + <FANCY_STUFF>.

**QUESTION: Recreate the plots below (pay attention to the details!)**

```{r,echo=FALSE}
histogram <- ggplot(data=iris, aes(x=Sepal.Width))
histogram + geom_histogram(binwidth=0.2, color="black", aes(fill=Species)) + 
  xlab("Sepal Width") +  ylab("Frequency") + ggtitle("Histogram of Sepal Width")
```

```{r,echo=FALSE}
ggplot(data=iris) + 
  geom_density(mapping=aes(x=Sepal.Length,fill=Species,col=Species),alpha=0.5) + 
  ggtitle('Iris dataset: Smoothed Density of Sepal Length per Species')
```

```{r,echo=FALSE}
smooth <- ggplot(data=iris, aes(x=Sepal.Length, y=Sepal.Width, color=Species)) + 
  geom_point(aes(shape=Species), size=1.5) + xlab("Sepal Length") + ylab("Sepal Width") + 
  ggtitle("Scatterplot with smoothers")

# Linear model
smooth + geom_smooth(method="lm")
```

```{r,echo=FALSE}
facet <- ggplot(data=iris, aes(Sepal.Length, y=Sepal.Width, color=Species)) + 
  geom_point(aes(shape=Species), size=1.5) + geom_smooth(method="lm") +
  xlab("Sepal Length") + ylab("Sepal Width") + ggtitle("Faceting")

# Along rows
facet + facet_grid(. ~ Species)
```

## Basic Data Manipulation

Make sure you have the data manipulation cheat sheet ready, because it will make the following section a lot easier.

### Filter rows

**Question: Filter by species 'virginica' and then show the summary statistics per variable**

```{r,echo=FALSE}
df <- filter(iris, Species == 'virginica')
summary(df)
```

### Select colums

**Question: Keep all columns that start with 'Petal' and show those summary statistics**

```{r,echo=FALSE}
select(iris,starts_with('Petal')) %>% summary()
```

### Mutate data

**Question: Create a new variable called Petal.Width.Rank which ranks Petal.Width in ascending order. Then use these values to recreate the plot below (watch the details!)**

```{r,echo=FALSE}
df <- iris
df <- mutate(df, Petal.Width.Rank = dplyr::percent_rank(Petal.Width))

ggplot(df) + geom_jitter(aes(x=Petal.Width.Rank,y=Petal.Width,color=Species))
```

## Data exploration

Now you have learned the basics of data manipulation and visualization.
Furthermore, you have seen that these cheat sheets are very useful when you need to perform an operation for which you do not know how to code it by heart.
You will now extend this knowledge by learning how to apply these skills to data exploration the first step in any data science project.
Please go through the [data exploration chapter](http://r4ds.had.co.nz/exploratory-data-analysis.html) in the R for Data Science book. **Don't forget to try out the exercises, there are 20 in total**.

## Assignment

There is no hand-in assignment this week, but please do go through the data exploration chapter mentioned above.

